Master's Project Journal

### 2019-04-13 thesis/FPGABasedADPLLs.tex
39 pages, onto chapter 4 of 6

\section{Frequency Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption{Frequency divider \ac{RTL} diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divs} (b), the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

To determine the sign itself, which clock edge first had a rising edge must be recorded, and this can be done by setting a \ac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle however, it is apparent that all 4 combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. Careful design of the system can avoid problems arising from this. Examining the truth table in Table \ref{table:sr_tt}, the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Table \ref{table:sr_tt} also illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.
\begin{table}[ht]
	\begin{center}
		\begin{tabular}{cc|cc}           
			$S$&$R$&$Q$&$\overline{Q}$\\
			\hline
			0&0&$Q_{n-1}$&$\overline{Q}_{n-1}$\T\\
			0&1&0&1\T\\
			1&0&1&0\T\\
			1&1&\multicolumn{2}{c}{undefined}\T\\					
		\end{tabular}
		\caption{SR Latch Truth Table}
		\label{table:sr_tt}
	\end{center}
\end{table}

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may in other systems be feasible to dismiss this occurrence as a low probability event, the goal of a \ac{PLL} is to reduce the time delay between rising edges to zero which increases the possibility of this occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in an \ac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is however a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on an \ac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on an \ac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
\begin{table}[!ht]
	\begin{center}
		\begin{tabular}{l|cc|cc|c|cc|c}           
			When&$Gen$&$Ref$&$\overline{Gen}/S^a$&$\overline{Ref}/R^a$&$Q^a$&$S^b$&$R^b$&$Sign/Q^b$\\
			\hline
			Between Measurements&0&0&1&1&\multicolumn{1}{c|}{undefined}&\multicolumn{2}{c|}{01 or 10}&\multicolumn{1}{c}{0 or 1}\T\\
			Ref. First&0&1&1&0&1&0&1&0\T\\
			Gen. First&1&0&0&1&0&1&0&1\T\\	
			Measurement Complete&1&1&0&0&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$\T\\					
		\end{tabular}
		\caption{Sign Detector Truth Table}
		\label{table:sign_tt}
	\end{center}
\end{table}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_sign_detection}
	\caption[Example Sign Detector]{Example Sign Detector.}
	\label{fig:arbitration}
\end{figure}

@@

As the scaling factor used in this multiplier is the $k_i$ gain itself, reconfiguration of the integral path of the \acl{LF} can be carried out solely by varying this gain, as opposed to the \ac{FIR} solution where the value of each tap must be recalculated to change the response of the system. The combination of reconfigurability and simpler hardware make the \ac{IIR} the more desirable \ac{LF} design.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi.pdf}
	\caption[Basic \ac{IIR} \ac{PI} loop filter]{Basic \ac{IIR} \ac{PI} loop filter.}
	\label{fig:iir_pi}
\end{figure}

Again both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring logic combining the two paths to preserve the relationship between the proportional and integral gains, which from the work of Koskin \textit{et al} is known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the filter becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between $k_p$ and $k_i$ is preserved during the combination of proportional and integral paths as both values have their radix point aligned before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}

>>>>

\section{Frequency Divider}
Many \ac{PLL}s, be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on an \ac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption[Frequency Divider \ac{RTL} Diagrams]{Frequency Divider \ac{RTL} Diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divs} (b), the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

@@

To determine the sign itself, which clock edge first had a rising edge must be recorded, and this can be done by setting a \ac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle however, it is apparent that all 4 combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. Careful design of the system can avoid problems arising from this. Examining the truth table in Table \ref{table:sr_tt}, the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Table \ref{table:sr_tt} also illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.
\begin{table}[ht]
	\begin{center}
		\begin{tabular}{cc|cc}           
			$S$&$R$&$Q$&$\overline{Q}$\\
			\hline
			0&0&$Q_{n-1}$&$\overline{Q}_{n-1}$\T\\
			0&1&0&1\T\\
			1&0&1&0\T\\
			1&1&\multicolumn{2}{c}{undefined}\T\\					
		\end{tabular}
		\caption{SR latch truth table}
		\label{table:sr_tt}
	\end{center}
\end{table}

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may in other systems be feasible to dismiss this occurrence as a low probability event, the goal of a \ac{PLL} is to reduce the time delay between rising edges to zero which increases the possibility of this occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in an \ac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is however a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on an \ac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on an \ac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
\begin{table}[!ht]
	\begin{center}
		\setlength{\tabcolsep}{.5\tabcolsep}
		\begin{tabular}{l|cc|cc|c|cc|c}           
			When&$Gen$&$Ref$&$\overline{Gen}/S^a$&$\overline{Ref}/R^a$&$Q^a$&$S^b$&$R^b$&$Sign/Q^b$\\
			\hline
			Between Measurements&0&0&1&1&\multicolumn{1}{c|}{undefined}&\multicolumn{2}{c|}{01 or 10}&\multicolumn{1}{c}{0 or 1}\T\\
			Ref. First&0&1&1&0&1&0&1&0\T\\
			Gen. First&1&0&0&1&0&1&0&1\T\\	
			Measurement Complete&1&1&0&0&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$\T\\					
		\end{tabular}
		\caption{Sign detector truth table}
		\label{table:sign_tt}
	\end{center}
\end{table}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_sign_detection}
	\caption[Example Sign Detector]{Example Sign Detector.}
	\label{fig:arbitration}
\end{figure}

@@

As the scaling factor used in this multiplier is the \ac{ki} itself, reconfiguration of the integral path of the \acl{LF} can be carried out solely by varying this gain, as opposed to the \ac{FIR} solution where the value of each tap must be recalculated to change the response of the system. The combination of reconfigurability and simpler hardware make the \ac{IIR} the more desirable \ac{LF} design.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi.pdf}
	\caption[Basic \ac{IIR} \ac{PI} loop filter]{Basic \ac{IIR} \ac{PI} loop filter.}
	\label{fig:iir_pi}
\end{figure}

Again both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring logic combining the two paths to preserve the relationship between the proportional and integral gains, which from the work of Koskin \textit{et al} is known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the \ac{ADPLL} becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between \ac{kp} and \ac{ki} is preserved during the combination of proportional and integral paths as both values have their radix point aligned before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}
